Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
regularized parens
node_contents

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}
	&\mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, (\mathsf{k}, \mathsf{v}), range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.

@@

The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$, indicating that a thread should already hold \lstinline{n}'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state. The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, (\mathsf{k'}, \mathsf{v'}), \mathsf{range'})
\end{align*}

@@

Figure \ref{proof_lock_traverse} shows the proof outline of the \texttt{traverse} function.
We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$. Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 9), which returns a new node \lstinline{n''} to visit, stored in \lstinline{pn-> n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$. Otherwise, we acquire \lstinline{n->lock}, gaining access to the resources $\mathsf{R}(\texttt{n})$, and then release the lock of the current node \texttt{p} and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
The two \lstinline{return}s are also the two possible linearization points of the function.

@@

The insert function uses this specification of \lstinline{traverse} to update the state of the data structure. The atomic specification for \texttt{insert} is
\begin{mathpar}
	{\color{blue}
		m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We can then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
We can then release the lock held by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

\wm{Should we get rid of inFP and just talk about Node\_ref?}
The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup}, is slightly simpler than that of the lock-coupling template: a thread does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition.
As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$ \wm{or node\_contents?}. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.

%\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly. 
For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held. %Formally, we write $$\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b. \ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}$$
%\wm{Consider using mathsf for logic and reserving texttt for C code.}
%where a lock at location $\ell$ safeguards resources denoted by \texttt{R}. When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$. On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
%and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.
As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \texttt{min} and \texttt{max} fields that we use to implement the \texttt{inRange} check.

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p r n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n'})
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(\texttt{r})$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

\section{Verifying Binary Search Trees using Templates}
\label{BST_proof}
In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. The top-level specifications we obtain for search structure operations are shown below: each operation takes effect on the data structure atomically, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
\begin{mathpar}
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\\
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ 
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\\
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
As we saw in the previous section, these top-level specifications have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g. \lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs. Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.

@@

\begin{figure}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function \\\\}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function \\\\}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure}

>>>>

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}
	&\mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, \mathsf{c}, range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{c})\\
	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{c}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $\mathsf{c}$ of a node can be either a key-value pair $(\mathsf{k}, \mathsf{v})$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).

@@

The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$, indicating that a thread should already hold \lstinline{n}'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}. The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n}, \mathsf{c}, \mathsf{range}.\ &\texttt{pn} \mapsto (\texttt{p, n}) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(\texttt{n})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n}, \mathsf{c}, \mathsf{range})
\end{align*}

@@

Figure \ref{proof_lock_traverse} shows the proof outline of the \texttt{traverse} function.
We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$. Otherwise, we pass the $\mathsf{node\_data}$ from the lock invariant to \lstinline{findNext} (line 11), which returns a new node \lstinline{n''} to visit, stored in \lstinline{pn-> n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$. Otherwise, we acquire \lstinline{n->lock}, gaining access to the resources $\mathsf{R}(\texttt{n})$, and then release the lock of the current node \texttt{p} and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.
The two \lstinline{return}s are also the two possible linearization points of the function.

@@

The insert function uses this specification of \lstinline{traverse} to update the state of the data structure. The atomic specification for \texttt{insert} is
\begin{mathpar}
	{\color{blue}
		m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We can then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
We can then release the lock held by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup}, is slightly simpler than that of the lock-coupling template: a thread does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition.
As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether \lstinline{n'} is empty or contains the key \lstinline{k}.

%\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly. 
%For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held. %Formally, we write $$\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists \ b. \ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}$$
%\wm{Consider using mathsf for logic and reserving texttt for C code.}
%where a lock at location $\ell$ safeguards resources denoted by \texttt{R}. When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$. On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
%and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure.

%As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \texttt{min} and \texttt{max} fields that we use to implement the \texttt{inRange} check.

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p}, \texttt{r}, \texttt{n}.\ \texttt{pn} \mapsto (\texttt{p, n}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n})
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(\texttt{r})$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

\section{Verifying Binary Search Trees using Templates}
\label{BST_proof}
In this section, we demonstrate how to instantiate the templates with specific data structures, giving us verified C implementations of concurrent data structures. The top-level specifications we obtain for search structure operations are shown below: each operation takes effect on the data structure atomically, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.
\begin{mathpar}
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\\
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ 
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\\
	{\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
As we saw in the previous section, these top-level specifications have already been verified at the template level; all we need to do now is provide implementations of data-structure-specific functions (e.g. \lstinline{findNext}) that satisfy the right specifications to instantiate the template proofs. Once we have chosen a template and a data structure, we should not need to do any further reasoning about specific concurrency patterns.

@@

\begin{figure}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function \\\\}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function \\\\}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure}
